# Data manipulation and plotting


## Setup
Cheatsheets for dplyr and ggplot2 are available in rstudio by using the Help menu (Help -> Cheatsheets)

Install and load the libraries
```{r}
# if libraries are not installed uncomment the next 3 lines
# install.packages('dplyr')
# install.packages('ggplot2')
# install.packages('tidyr')

library(dplyr)
library(ggplot2)
library(tidyr)
```

## Data manipulation using dplyr

Have a look at the iris dataset and introduce %>% for piping

Most commands will end with a '%>% head()' to reduce excessive results being displayed
```{r}
data("iris")
View(iris)

iris %>% head()
head(iris)
names(iris)
colnames(iris)
```


Demonstrate standard versus dplyr method for selecting a column 
```{r}
iris[,"Sepal.Length"]
iris$Sepal.Length
# select 2 columns with dplyr
iris %>% select(Sepal.Length, Sepal.Width) %>% head()
# use -columnname to deselect that column
iris %>% select( -Species) %>% head()
```


Choose rows using filter()

```{r}
iris %>% select(Sepal.Length, Sepal.Width) %>% filter(Sepal.Length > 5) %>% head()
```

Summarise data
```{r}
iris %>% 
  select(Sepal.Length, Sepal.Width) %>% 
  filter(Sepal.Length > 5) %>% 
  summarise(min(Sepal.Length), max(Sepal.Width))

iris %>% 
  select(Sepal.Length, Sepal.Width) %>% 
  filter(Sepal.Length > 5) %>% 
  summarise(minLength = min(Sepal.Length), maxWidth = max(Sepal.Width))
```


Use of group_by() to summarise by groups
```{r}
iris %>% 
  select(Sepal.Length, Sepal.Width, Species) %>% 
  filter(Sepal.Length > 5) %>% 
  group_by(Species) %>% 
  summarise(minLength = min(Sepal.Length), maxWidth = max(Sepal.Width))
```

tally() to count rows and compare with results from dim()
```{r}
iris %>% 
  select(Sepal.Length, Sepal.Width, Species) %>% 
  filter(Sepal.Length > 5) %>% 
  tally()

dim(iris[iris$Sepal.Length > 5, ])
```

tally() on groups
```{r}
iris %>% 
  select(Sepal.Length, Sepal.Width, Species) %>% 
  filter(Sepal.Length > 5) %>% 
  group_by(Species) %>% 
  tally()
```


creating new columns with mutate()
```{r}
iris %>% 
  select(Sepal.Length, Sepal.Width, Species) %>% 
  filter(Sepal.Length > 5) %>% 
  mutate(Length100 = Sepal.Length * 100, length100_2 = Length100 * 2) %>% 
  head()
```


sorting data in ascending order based on column using arrange()
```{r}
iris %>% 
  select(Sepal.Length, Sepal.Width, Species) %>% 
  filter(Sepal.Length > 5) %>% 
  arrange(Sepal.Length) %>% 
  head()
# arranged in descending order using desc() inside arrange()
iris %>% 
  select(Sepal.Length, Sepal.Width, Species) %>% 
  filter(Sepal.Length > 5) %>% 
  arrange(desc(Sepal.Length)) %>% 
  head()
```


Showing the unique rows with distinct()
```{r}
iris %>% 
  select(Sepal.Length, Species) %>% 
  filter(Sepal.Length > 5) %>% 
  arrange(desc(Sepal.Length)) %>% 
  select(Sepal.Length, Species) %>% 
  distinct() %>% 
  head()

#tally distinct rows
iris %>% 
  select(Sepal.Length, Species) %>% 
  filter(Sepal.Length > 5) %>% 
  arrange(desc(Sepal.Length)) %>% 
  select(Sepal.Length, Species) %>% 
  distinct() %>% 
  tally()

# tally without distinct()
iris %>% 
  select(Sepal.Length, Species) %>% 
  filter(Sepal.Length > 5) %>% 
  arrange(desc(Sepal.Length)) %>% 
  select(Sepal.Length, Species) %>% 
  tally()

```





## ggplot2

### documentation for ggplot2: http://docs.ggplot2.org/current/

ggplot2 uses a method of mapping the data, defining the aesthetics, and finally using geometric obects to represent the data


base plot example
```{r}
plot(iris$Sepal.Length, iris$Width)
```


scatter plot
```{r}
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width)) + geom_point()
```

histogram

```{r}
ggplot(iris, aes(x = Sepal.Length)) + geom_histogram()
```

line plot
```{r}
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width)) + geom_line()
```



box plot
```{r}
# all data
ggplot(iris, aes(x = 1, y = Sepal.Length)) + geom_boxplot()
# separate boxes for each species
ggplot(iris, aes(x = Species, y = Sepal.Length)) + geom_boxplot()

```


Demonstrated in this command
- plot being saved as a object
- multiple geoms being used
- changing plot and axis titles
- changing colours
- using a different theme
```{r}

p <- ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, colour = Species, shape = Species)) + 
  geom_point() + 
  geom_line(aes(colour = "red")) + 
  ylab("Width") + xlab("Length (mm)") + 
  ggtitle("My Awesome Plot") + 
  theme_bw()

p
```

using ggsave() to save a plot
```{r, eval=FALSE}
ggsave(filename = '~/pretty_plot.png', plot = p, device = 'png')    
```

using the last plot function to replot the last plot with no legend
```{r}
last_plot() + theme(legend.position = 'none')
```



showing dplyr being combined with ggplot
```{r}
iris %>% 
  select(Sepal.Length, Sepal.Width, Species) %>% 
  filter(Sepal.Length > 5) %>% 
  ggplot(., aes(x = Sepal.Length, y =Sepal.Width)) + geom_point()
```



tidyr example attempt to collapse sepal.length and sepal.width into a grouping variable (dimension) and a data variable (mm)

```{r}
iris %>% 
  select(Species, Sepal.Length, Sepal.Width) %>% 
  gather(.,"dimension", "mm", 2:3) %>% # gather is from tidyr
  ggplot(., aes(x = dimension, y = mm, fill = Species)) + 
  geom_bar(stat = 'identity', position = 'dodge')
```


A recreation of the SWC loops lesson for plotting multiple datasets
```{r}
# create the plotting function
plot_graphs <- function(df){
  p <- ggplot(df, aes(x = Sepal.Length, y = Sepal.Width)) + geom_point()
  return(p)
}

# create some example datasets
df1 <- iris %>% filter(Species == 'setosa')
df2 <- iris %>% filter(Species == 'virginica')
datasets <- list(df1, df2)

# plot using a loop
for(d in datasets){
  plot(plot_graphs(d))
}

```





